The zero--one laws of Kolmogorov and Hewitt--Savage in categorical probability

### 2019-11-23 categorical-zero-one.tex
another revision

\begin{abstract}
	We state and prove the zero--one laws of Kolmogorov and Hewitt--Savage within the setting of \emph{Markov categories}, a category-theoretic approach to the foundations of probability and statistics. This gives general versions of these results which can be instantiated not only in measure-theoretic probability, where they specialize to the standard ones in the setting of standard Borel spaces, but also in other contexts. For example, applying them to the Kleisli category of the hyperspace monad on topological spaces gives criteria for when maps out of an infinite product of topological spaces are constant.
\end{abstract}

@@

\emph{Markov categories} are an approach to probability and statistics based on category theory, developed by the first author in \cite{markov_cats}, building on much previous work by other authors (see \Cref{background}).
The basic idea is to axiomatize the structure and properties of ``categories of stochastic processes'' - for instance, the category $\Stoch$ where objects are measurable spaces and morphisms are Markov kernels (hence the name).
The goal is to state and prove results from probability and statistics in this context.
This represents a \emph{synthetic} approach to probability, in contrast with the usual measure-theoretic approach (which can be called \emph{analytic}).
Instead of defining \emph{what} the words probability, distribution, random variable, etc, \emph{mean}, we instead state by which operations these objects can be combined and related to each other, and which properties they are supposed to satisfy.
We can draw an analogy to the Euclidean approach to geometry - where points, lines and so on are described only in terms of their relationships with one another - with the more modern approach, where geometric objects are defined concretely in terms of more primitive notions (i.e sets, real numbers, etc).

This is a treatment of the zero--one laws of Kolmogorov and of Hewitt--Savage in the setting of Markov categories.

Very roughly, the Kolmogorov zero--one law states that, given an infinite collection of independent random variables, and an event which is in the $\sigma$-algebra generated by the variables, which is independent of any finite subset of variables, the event has probability one or zero.
It goes back to Kolmogorov's foundational monograph \cite{kolmogorovgerm} - although the statement appearing there is significantly different from the modern form.
For a more recent treatment of this theorem, see e.g. \cite[Theorem~2.37]{klenke}

The Hewitt--Savage zero--one law, due to Hewitt and Savage in \cite[Theorem~11.3]{hewitt_savage} is similar - here the variables are supposed to be independent and indetically distributed, and the event is supposed to be invariant under finite permutations of the variables. Again, under these assumptions, the event has probability one or zero.

To formulate these statements in the general context of Markov categories, we need a notion of infinite family of random variables.
To this end, we define a notion of \emph{infinite tensor product}. Given such a tensor product, satisfying a certain condition of compatibility with the structure of the Markov category (\Cref{defn_kolmogorov_ext}), we prove abstract versions of the zero--one laws, \Cref{thm:kolmog} and \Cref{thm:hewsav}.

Last, we apply our results to the case of measurable spaces and Markov kernels, deriving the zero--one laws for random variables in standard Borel spaces (this regularity assumption is needed to make the necessary infinite tensor products exist).

@@

In \Cref{infprod_semicartesian}, $\cC$ denotes more concretely a semicartesian strict symmetric monoidal category. In \Cref{infprod_markov} and after, $\cC$ denotes a Markov category in the sense of \Cref{markov_cat}. $J$ is a set used for indexing products of objects; the definitions and results of this paper are nontrivial only when $J$ is infinite. $F$ denotes either an arbitrary finite set or more concretely a finite subset of $J$.

\paragraph*{Acknowledgments.} We thank Paolo Perrone for discussions leading to \Cref{poset_case}.

@@

A symmetric monoidal category $\cC$ is \emph{semicartesian} if the monoidal unit $I \in \cC$ is terminal. Equivalently\todo{Should there be a reference to this equivalence?}, it is a monoidal category equipped with morphisms
\[
	X \otimes Y \longrightarrow X, \qquad X \otimes Y \longrightarrow Y,
\]
which are natural in $X$ and $Y$ and coincide with the monoidal structure isomorphisms whenever $X = I$ or $Y = I$. Here are the main examples that we will be considering.

\begin{example}
	\label{finstoch}
	$\FinStoch$ is the category of finite sets with \emph{stochastic matrices}, which means that $f : X \to Y$ is a matrix $(f_{xy})_{x \in X,y \in Y}$ of nonnegative real numbers such that $\sum_y f_{xy} = 1$ for every $x$. Stochastic matrices compose by matrix multiplication, which in this context is also known as the \emph{Chapman--Kolmogorov equation}. We consider $\FinStoch$ as a symmetric monoidal category with respect to the cartesian product of sets on objects, and as one would expect the Kronecker product of stochastic matrices on morphisms.
\end{example}

@@

\begin{example}
	\label{stoch}
	$\Stoch$ is the category of measurable spaces $(X,\Sigma_X)$ as objects and Markov kernels as morphisms. We sketch the definition here and refer to~\cite[Section~4]{markov_cats} and references therein for the details. A \emph{Markov kernel} $(X,\Sigma_X) \to (Y,\Sigma_Y)$ is a function
	\[
		f \: : \: \Sigma_Y \times X \longrightarrow [0,1], \quad (S,x) \longmapsto f(S|x)
	\]
	assigning to every $x \in X$ a probability measure $f(-|x) : \Sigma_Y \to [0,1]$ in such a way that for every $S \in \Sigma_Y$, the function $x \mapsto f(S|x)$ is measurable. Intuitively, $f$ assigns to every $x \in X$ a random element of $Y$. Composition is again defined by a variant of the Chapman--Kolmogorov equation, and the symmetric monoidal structure is defined in terms of the usual product of measurable spaces. The monoidal unit $I$ is given by any one-element set with its unique $\sigma$-algebra. As a special case, morphisms $I \to (X,\Sigma_X)$ can be identified with probability measures $\sigma_X \to [0,1]$.

	There is a well-known subclass of particularly well-behaved measurable spaces, the \emph{standard Borel spaces} or equivalently Polish spaces equipped with their Borel $\sigma$-algebras. We write $\BorelStoch \subseteq \Stoch$ for the full subcategory of standard Borel spaces with Markov kernels. Since finite products of standard Borel spaces are again standard Borel, $\BorelStoch$ is again semicartesian symmetric monoidal.
\end{example}

Returning to the general theory, in string diagrams the unique morphism $\discard{X} : X \to I$ for an object $X \in \cC$ is denoted by
\[
	\tikzfig{terminal}
\]
The projection maps $\id \otimes \discard{Y} : X \otimes Y \to X$ and $\discard{Y} \otimes \id : X \otimes Y \to Y$ are correspondingly written as
\[
	\tikzfig{marginals}
\]

Our goal is to use semicartesian monoidal categories such as the above in order to develop aspects of probability theory in categorical terms. As it turns out, doing so requires a bit more structure, in a form which has been axiomatized first by Cho and Jacobs~\cite{cho_jacobs} as \emph{affine CD-categories}, although similar definitions occur in earlier work of Golubtsov~\cite{golubtsov}. We here follow the terminology of our own~\cite[Definition~2.1]{markov_cats}.

\begin{definition}
	A \emph{Markov category} $\cC$ is a semicartesian symmetric monoidal category where every object $X \in \cC$ is equipped with a morphism
	\begin{equation}
		\label{comultiplication}
		\tikzfig{comultiplication}
	\end{equation}
	which, together with $\discard{X} : X \to I$, makes $X$ into a commutative comonoid, and such that
	\begin{equation}
		\label{multiplicativity}
		\tikzfig{multiplicativity}
	\end{equation}
	for all $X, Y \in \cC$.
	\label{markov_cat}
\end{definition}

@@

A closely related definition was also used in earlier work of Golubtsov~\cite{golubtsov}, who had proposed a closely related definition, and Cho and Jacobs~\cite{cho_jacobs}, where an equivalent definition was used under the term \emph{affine CD-category}.

We think of the comultiplication~\Cref{comultiplication} as a copying operation. The coassociativity and counitality conditions guarantee that copying with any number of output wires is well-defined, and we draw it likewise as a single black dot with any number of outgoing wires, as in
\[
	\tikzfig{triple_copy}
\]

@@

As per~\cite[Remark~10.12]{markov_cats}, the deterministic morphisms form a symmetric monoidal cartesian subcategory $\cC_\detc \subseteq \cC$ which contains all structure morphisms of $\cC$, including the comultiplications themselves. The fact that all morphisms are comonoid homomorphisms implies that $\cC_\detc$ is cartesian monoidal.

In $\FinStoch$, the deterministic morphisms $X \to Y$ are exactly the $\{0,1\}$-valued stochastic matrices. Since these can be identified with honest functions $X \to Y$, we conclude that $\FinStoch_\detc$ is equivalent to the category of finite sets. In $\Stoch$, the deterministic morphisms $f : X \to Y$ are exactly those Markov kernels for which $f(S|x) \in \{0,1\}$ for every $S \in \Sigma_Y$ and $x \in X$. In other words, as the term suggests, the deterministic morphisms are those which do not involve any randomness. In $\CRing_+\op$, the deterministic morphisms $R \to S$ are precisely those morphisms which are represented by multiplicative unital maps $S \to R$, i.e.~the ring homomorphisms. Thus the subcategory of deterministic morphisms is exactly the opposite of the usual category of commutative rings.

@@

If $\bigotimes_{j \in F} X_j$ is a finite product without any particular order on the factors\footnote{See e.g.~\cite[Proposition~II.1.5]{DMOS} for how to make sense of tensor products in symmetric monoidal categories without a prescribed order of the factors.}, then we also write $\perp_{j \in F} X_j \mid\mid A$ to denote the corresponding conditional independence for given $p : A \to \bigotimes_{j \in F} X_j$.

\section{Infinite tensor products in semicartesian monoidal categories}
\label{infprod_semicartesian}

@@

\begin{definition}
	\label{semicartesian_infproduct}
	Let $(X_j)_{j \in J}$ be a family of objects in $\cC$. An \emph{infinite tensor product}
	\[
		X_J \: := \: \bigotimes_{j \in J} X_j
	\]
	is the limit of the diagram $F \mapsto X_F$, indexed by the poset of finite subsets $F \subset J$ ordered by reverse inclusion, if this limit exists and is preserved by the functor $- \otimes Y$ for every object $Y \in \cC$.
\end{definition}

@@

\begin{remark}
	\label{preserve}
	It is natural to require different infinite tensor products to interact well with another. If $X_J$ is an infinite tensor product of a family $(X_i)_{i \in J}$ and we are given an additional object $X_\ast$ for $\ast \not \in J$, then one will want the canonical comparison morphism
	\begin{equation}
		\label{comparison}
		X_J \otimes X_\ast \longrightarrow X_{J \,\sqcup\, \{\ast\}}
	\end{equation}
	to be an isomorphism, where $X_{J \,\sqcup\, \{\ast\}}$ is the infinite tensor product of the original family with $X_\ast$ thrown in.

	If we do not yet assume that $X_{J \,\sqcup\, \{\ast\}}$ exists, then we can exhibit $X_J \otimes X_\ast$ as the relevant cofiltered limit as follows. The object $X_J \otimes X_\ast$ can be equipped finite marginalization morphisms with respect to $F \subseteq J \sqcup \{\ast\}$ given by
	\[
		\pi_{F \setminus \{\ast\}} \otimes \id \: : \: X_J \otimes X_\ast \longrightarrow X_{F \setminus \{\ast\}} \otimes X_\ast
	\]
	for $\ast \in F$, and just
	\[
		\pi_F \otimes \discard{X_\ast} : X_J \otimes X_\ast \longrightarrow X_F 
	\]
	in case $\ast \not \in F$. It is straightforward to check that these morphisms equip $X_J \otimes X_\ast$ with the right universal property of the infinite tensor product $X_{J \sqcup \{\ast\}}$ if and only if the functor $- \otimes X_\ast$ preserves the defining cofiltered limit of the infinite tensor product $X_J$. It is also easy to see that this new cofiltered limit is again preserved by every functor $- \otimes Y$, based on the assumption that the defining cofiltered limit of the original infinite tensor product is preserved by $- \otimes (X_\ast \otimes Y)$ as well.

	In other words, the preservation condition in this definition amounts to the requirement that the canonical comparison morphism
	must be an isomorphism, as one would expect intuitively from a notion of infinite tensor product. 
\end{remark}

\begin{example}
	\label{infprods_cring}
	The dual definition of infinite tensor products of algebraic structures as \emph{filtered colimits} of finite tensor products is well-known in the literature, e.g.~in the case of C*-algebras~\cite[p.~315]{blackadar}. Thus in the case where our Markov category is $\CRing_+\op$, we recover the usual folklore definition of an infinite tensor product of rings $\bigotimes_{j \in J} R_j$ in terms of formal sums of elementary tensors, where an elementary tensor is a family of elements $(r_j)_{j \in J}$ such that all but finitely many are equal to the respective unit. The necessary preservation condition is easily seen to hold. Note that we do not yet consider the multiplication on $\bigotimes_{j \in J} R_j$, since \Cref{semicartesian_infproduct} is not yet concerned with the comonoid structures on the objects. We will get to this in the next section.

	Intuitively, our definition matches up with the known algebraic ones under the categorical duality of algebra and geometry, where our definition is on the geometrical side of the duality.
\end{example}

\begin{remark}
	In those semicartesian monoidal categories $\cC$ that are of interest to us, morphisms $I \to X$ play the role of \emph{probability measures} on $X$. Thus applying the defining universal property of an infinite tensor product with respect to maps out of $I$ implements the \emph{Kolmogorov extension theorem} in $\cC$: probability measures on an infinite product $X_J$ are in bijection with consistent families of probability measures on the finite products $X_F$ for $F \subseteq J$.
\end{remark}

@@

	\label{infprods_stoch}
	We now consider infinite tensor products in $\Stoch$, where we would like infinite tensor products to be given by the corresponding infinite products of measurable spaces in the usual sense. Since the Kolmogorov extension theorem does not hold for general (even merely countable) products of measurable spaces~\cite{AJ}, this is not the case without further additional assumptions on the measurable spaces involved. Thus we do not know whether $\Stoch$ has infinite tensor products, although we suspect that it does not; but even if it does, they are in general not the ones that one would like to have for the purposes of probability theory.

	However, the situation improves for countable products in $\BorelStoch$, for which the Kolmogorov extension theorem holds~\cite[Theorem~14.35]{klenke}. In other words, if $\left( (X_j, \Sigma_j) \right)_{j \in \bN}$ is a sequence of standard Borel spaces, then the cartesian product $X_\bN = \prod_{j \in \bN} X_j$ carrying the product $\sigma$-algebra $\Sigma_\bN$ satisfies the universal property of an infinite tensor product with respect to maps out of $I$. We now show that this implies the universal property in general.

	Suppose that
	\[
		\Big( g_F : (A,\Sigma_A) \longrightarrow (X_F, \Sigma_F) \Big)_{F \subseteq J \text{ finite}}
	\]
	is a family of Markov kernels satisfying the compatibility condition $\pi_{F',F} \circ g_{F'} = g_F$ for all finite $F \subseteq F' \subseteq J$. Then for every $a \in A$, the probability measures $g_F(-|a) : \Sigma_F \to [0,1]$ are a compatible family to which the Kolmogorov extension theorem in the form~\cite[Theorem~14.35]{klenke} applies, and we obtain a unique probability measure $g_J(-|a) : \Sigma_J \to [0,1]$ which has the $g_F(-|a)$ as its finite marginals. It remains to be shown that for every $S \in \Sigma_J$, the map $a \mapsto g_J(S|a)$ is measurable. Since limits of pointwise convergent sequences of measurable real-valued functions are again measurable, the set of $S$ for which this measurability holds is closed under countable disjoint union, and it is clearly closed under complements. We therefore have a $\lambda$-system. Since the map is measurable by assumption whenever $S$ is a measurable cylinder set, and the cylinder sets form a $\pi$-system, the $\pi$-$\lambda$-theorem implies that $a \mapsto g_J(S|a)$ is measurable for all $S$ in the $\sigma$-algebra generated by the measurable cylinder sets, which is exactly the product $\sigma$-algebra $\Sigma_J$.

@@

\begin{definition}
    \label{defn_kolmogorov_ext}
    Let $\cC$ be a Markov category and $(X_j)_{j \in J}$ a family of objects. We say that an infinite tensor product $X_J = \bigotimes_{j \in J} X_j$ is a \emph{Kolmogorov product} if the finite marginalization morphisms $\pi_F : X_J \to X_F$ are deterministic.
\end{definition}

In $\BorelStoch$, the infinite tensor products constructed in \Cref{infprods_stoch} are Kolmogorov products. In $\CRing_+\op$, the infinite tensor products from \Cref{infprods_cring} are Kolmogorov products as well as long as one equips $\bigotimes_j R_j$ with the tensor product ring structure, since this one is the only one which makes the canonical inclusions $R_j \to \bigotimes_j R_j$ into ring homomorphisms. These examples illustrate the following important point. In some Markov categories, such as $\CRing_+\op$, not all isomorphisms are deterministic, and there may even be isomorphic objects which have no deterministic isomorphism~\cite[Remark~10.9]{markov_cats}. In these categories, whether a given infinite tensor product is a Kolmogorov product or not generally depends on the specific choice of that object. In particular, there may be infinite tensor products which are not Kolmogorov products. For example in $\CRing_+\op$, we may take the tensor product of abelian groups $\bigotimes_j R_j$ and equip it with a different multiplication than the tensor product of rings one (as long as it has the same unit).

>>>>

\begin{abstract}
	We state and prove the zero--one laws of Kolmogorov and Hewitt--Savage within the setting of \emph{Markov categories}, a category-theoretic approach to the foundations of probability and statistics. This gives general versions of these results which can be instantiated not only in measure-theoretic probability, where they specialize to the standard ones in the setting of standard Borel spaces, but also in other contexts. For example, applying them to the Kleisli category of the hyperspace monad on topological spaces gives criteria for when maps out of an infinite product of topological spaces into a Hausdorff space are constant.
\end{abstract}

@@

\emph{Markov categories} are an approach to the foundations of probability and statistics based on category theory, proposed first by Golubtsov~\cite{golubtsov}, rediscovered recently independently by Cho and Jacobs~\cite{cho_jacobs}, and developed extensively by the first author in \cite{markov_cats}. The basic observation is that Markov kernels can be composed sequentially and in parallel, making them into a symmetric monoidal category $\Stoch$. One can then find axioms which make a symmetric monoidal category behave similarly to the actual category of Markov kernels, and then state and prove results from probability and statistics in this general abstract context.

This represents a \emph{synthetic} approach to probability, in contrast with the usual measure-theoretic approach (which can be called \emph{analytic}). Instead of defining \emph{what} the words probability, distribution, random variable, etc, \emph{mean}, we instead state by which operations these objects can be combined and related to each other, and which properties they are supposed to satisfy.
We can draw an analogy to the Euclidean approach to geometry: here, points and lines are described only in terms of their relationships with one another; while in the analytic approach often attributed to Descartes, where geometric objects are defined concretely in terms of more primitive notions like sets and real numbers.

In this paper, we give a treatment of the zero--one laws of Kolmogorov and of Hewitt--Savage in the setting of Markov categories. This adds further support to the hypothesis that Markov categories are an adequate setting for the foundations of probability and statistics, adding to the existing treatments of Bayesian updating, almost sure equality, conditional independence, and theorems on sufficient statistics~\cite{cho_jacobs,markov_cats}.

Very roughly, the Kolmogorov zero--one law states that, given an infinite collection of independent random variables and an event which is in the $\sigma$-algebra generated by the variables but independent of any finite subset of these variables, the event has probability zero or one.
This theorem goes back to Kolmogorov's foundational monograph \cite{kolmogorovgerm}, although the statement appearing there is significantly different from the modern form.
For a more recent treatment of this theorem, see e.g. \cite[Theorem~2.37]{klenke}

The Hewitt--Savage zero--one law, due to Hewitt and Savage in \cite[Theorem~11.3]{hewitt_savage}, is similar. Here, the variables are assumed to be identically distributed in addition to being independent, and the event is assumed to be invariant under finite permutations of the variables. Again, under these assumptions, the event has probability zero or one.

To formulate these statements in the general context of Markov categories, we need an abstract generalization of the concept of joint distribution of infinitely many random variables.
To this end, we introduce a notion of \emph{infinite tensor product}, first for semicartesian monoidal categories (\Cref{infprod_semicartesian}) and then for Markov categories (\Cref{infprod_markov}). A morphism to such an infinite tensor product is an abstract version of a probability measure on an infinite product of measurable spaces. Based on this definition, we then prove abstract versions of the zero--one laws, \Cref{thm:kolmog} and \Cref{thm:hewsav}.

Finally in \Cref{examples}, we apply our results to the category of measurable spaces and Markov kernels $\Stoch$. In order to guarantee that countably infinite tensor products exist, we restrict to the full subcategory of standard Borel spaces. We thereby rederive the standard zero--one laws for random variables in standard Borel spaces from our abstract theorems. We also instantiate our theorems in the Kleisli category of the lower Vietoris monad, which gives criteria for when a continuous map out of infinite product of topological spaces into a Hausdorff space is constant (\Cref{vietoris_kolmogorov} and \Cref{vietoris_hewitt_savage}).

@@

In \Cref{infprod_semicartesian}, $\cC$ denotes more concretely a semicartesian strict symmetric monoidal category. In \Cref{infprod_markov} and after, $\cC$ denotes a Markov category in the sense of \Cref{markov_cat}. $J$ is a set used for indexing products of objects; the definitions and results of this paper are nontrivial only when $J$ is infinite. $F$ denotes either an arbitrary finite set, or more concretely a finite subset of $J$.

\paragraph*{Acknowledgments.} We thank Paolo Perrone for discussions and \Cref{poset_case}.

@@

A symmetric monoidal category $\cC$ is \emph{semicartesian} if the monoidal unit $I \in \cC$ is terminal. Equivalently~\cite[Theorem~3.5]{GLS}, it is a symmetric monoidal category equipped with morphisms
\begin{equation}
	\label{projections}
	X \otimes Y \longrightarrow X, \qquad X \otimes Y \longrightarrow Y,
\end{equation}
which are natural in $X$ and $Y$ and coincide with the monoidal structure isomorphisms whenever $X = I$ or $Y = I$. Here are the main examples that we will be considering.

\begin{example}
	\label{finstoch}
	$\FinStoch$ is the category of finite sets with \emph{stochastic matrices}, which means that a morphism $f : X \to Y$ is a matrix $(f_{xy})_{x \in X,y \in Y}$ of nonnegative real numbers such that $\sum_y f_{xy} = 1$ for every $x$. Stochastic matrices compose by matrix multiplication, which in this context is also known as the \emph{Chapman--Kolmogorov equation}. We consider $\FinStoch$ as a symmetric monoidal category with respect to the cartesian product of sets on objects and the Kronecker product of stochastic matrices on morphisms; the monoidal structure isomorphisms are the obvious ones coming from the full embedding $\FinSet \subseteq \FinStoch$.
\end{example}

@@

\begin{example}
	\label{stoch}
	$\Stoch$ is the category of measurable spaces $(X,\Sigma_X)$ as objects and Markov kernels as morphisms. We sketch the definition here and refer to~\cite[Section~4]{markov_cats} and references therein for the details. A \emph{Markov kernel} $(X,\Sigma_X) \to (Y,\Sigma_Y)$ is a map
	\[
		f \: : \: \Sigma_Y \times X \longrightarrow [0,1], \quad (S,x) \longmapsto f(S|x)
	\]
	assigning to every $x \in X$ a probability measure $f(-|x) : \Sigma_Y \to [0,1]$ in such a way that for every $S \in \Sigma_Y$, the function $x \mapsto f(S|x)$ is measurable. Intuitively, $f$ assigns to every $x \in X$ a random element of $Y$. Composition is again defined by a variant of the Chapman--Kolmogorov equation, and the symmetric monoidal structure is defined in terms of the usual product of measurable spaces. The monoidal unit $I$ is given by any one-element set with its unique $\sigma$-algebra. As a special case, morphisms $I \to (X,\Sigma_X)$ can be identified with probability measures $\Sigma_X \to [0,1]$.

	There is a well-known subclass of particularly well-behaved measurable spaces, the \emph{standard Borel spaces}, or equivalently Polish spaces equipped with their Borel $\sigma$-algebras. We write $\BorelStoch \subseteq \Stoch$ for the full subcategory of standard Borel spaces with Markov kernels. Since finite products of standard Borel spaces are again standard Borel, $\BorelStoch$ is again semicartesian symmetric monoidal.
\end{example}

Returning to the general theory, in terms of string diagrams the unique morphism $\discard{X} : X \to I$ for an object $X \in \cC$ is denoted by
\[
	\tikzfig{terminal}
\]
The projection maps $\id \otimes \discard{Y} : X \otimes Y \to X$ and $\discard{Y} \otimes \id : X \otimes Y \to Y$, which are the ones from~\eqref{projections}, are correspondingly written as
\[
	\tikzfig{marginals}
\]

Our goal is to use semicartesian monoidal categories in order to develop aspects of probability theory in categorical terms, in such a way that instantiating this theory in $\Stoch$ or $\BorelStoch$ recovers the standard theory. As it turns out, doing so requires a bit more structure, in a form which has been axiomatized first by Cho and Jacobs~\cite{cho_jacobs} as \emph{affine CD-categories}, although very similar definitions occur in earlier work of Golubtsov~\cite{golubtsov}. We here follow the terminology of our own~\cite[Definition~2.1]{markov_cats}.

\begin{definition}
	A \emph{Markov category} $\cC$ is a semicartesian symmetric monoidal category where every object $X \in \cC$ is equipped with a distinguished morphism
	\begin{equation}
		\label{comultiplication}
		\tikzfig{comultiplication}
	\end{equation}
	which, together with $\discard{X} : X \to I$, makes $X$ into a commutative comonoid, and such that
	\begin{equation}
		\label{multiplicativity}
		\tikzfig{multiplicativity}
	\end{equation}
	for all $X, Y \in \cC$.
	\label{markov_cat}
\end{definition}

@@

We think of the comultiplication~\Cref{comultiplication} as a copying operation. The coassociativity and counitality conditions guarantee that copying with any number of output wires is well-defined, and we draw it likewise as a single black dot with any number of outgoing wires, like this:
\[
	\tikzfig{triple_copy}
\]

@@

As per~\cite[Remark~10.12]{markov_cats}, the deterministic morphisms form a symmetric monoidal subcategory $\cC_\detc \subseteq \cC$ which contains all structure morphisms of $\cC$, including the comultiplications themselves. The fact that all morphisms are comonoid homomorphisms implies that $\cC_\detc$ is actually cartesian monoidal.

In $\FinStoch$, the deterministic morphisms $X \to Y$ are exactly the $\{0,1\}$-valued stochastic matrices. Since these can be identified with honest functions $X \to Y$, we conclude that $\FinStoch_\detc$ is equivalent to $\FinSet$. In $\Stoch$, the deterministic morphisms $f : X \to Y$ are exactly those Markov kernels for which $f(S|x) \in \{0,1\}$ for every $S \in \Sigma_Y$ and $x \in X$. In other words, as the term suggests, the deterministic morphisms are those which do not involve any randomness; in $\BorelStoch$, the deterministic morphisms can be identified with the measurable functions. In $\CRing_+\op$, the deterministic morphisms $R \to S$ are precisely those morphisms which are represented by additive unital maps $S \to R$ which are in addition multiplicative, i.e.~the ring homomorphisms. Thus the subcategory of deterministic morphisms is exactly the opposite of the usual category of commutative rings.

@@

Conditional independence does not depend on the order of the tensor factors: if $p$ displays the conditional independence $X_1 \perp \ldots \perp X_n \mid\mid A$, then composing $p$ with any permutation $\sigma$ of the tensor factors $X_1,\ldots,X_n$ gives a morphism $A \to X_{\sigma(1)} \otimes \ldots \otimes X_{\sigma(n)}$ which displays the conditional independence $X_{\sigma(1)} \perp \ldots \perp X_{\sigma(n)} \mid\mid A$. Therefore if $\bigotimes_{j \in F} X_j$ is a finite product without any particular order on the factors\footnote{See e.g.~\cite[Proposition~II.1.5]{DMOS} for how to make sense of tensor products in symmetric monoidal categories without a prescribed order of the factors.}, then for $p : A \to \bigotimes_{j \in F} X_j$ there is no ambiguity about whether $p$ displays conditional independence or not, and we also write $\perp_{j \in F} X_j \mid\mid A$ in case that it does.

\section{Infinite tensor products in semicartesian symmetric monoidal categories}
\label{infprod_semicartesian}

@@

\begin{definition}
	\label{semicartesian_infproduct}
	Let $(X_i)_{i \in J}$ be a family of objects in $\cC$. An \emph{infinite tensor product}
	\[
		X_J \: := \: \bigotimes_{j \in J} X_j
	\]
	is the limit of the diagram $F \mapsto X_F$, indexed by the poset of finite subsets $F \subseteq J$ ordered by reverse inclusion, if this limit exists and is preserved by the functor $- \otimes Y$ for every object $Y \in \cC$.
\end{definition}

@@

\begin{remark}
	\label{preserve}
	It is natural to require different infinite tensor products to interact well with another. If $X_J$ is an infinite tensor product of a family $(X_i)_{i \in J}$ and we are given an additional object $X_\ast$ for $\ast \not \in J$, then one will want the canonical comparison morphism
	\begin{equation}
		\label{comparison}
		X_J \otimes X_\ast \longrightarrow X_{J \,\sqcup\, \{\ast\}}
	\end{equation}
	to be an isomorphism, where $X_{J \,\sqcup\, \{\ast\}}$ is the infinite tensor product of the original family with $X_\ast$ thrown in. We now show that this is indeed the case thanks to the preservation condition in \Cref{semicartesian_infproduct}, even if $X_{J \,\sqcup\, \{\ast\}}$ does not exist a priori.

	To do so, we exhibit $X_J \otimes X_\ast$ as the relevant cofiltered limit, which equips it with the defining universal property of $X_{J \,\sqcup\, \{\ast\}}$, as follows. The object $X_J \otimes X_\ast$ can be equipped with finite marginalization morphisms with respect to $F \subseteq J \sqcup \{\ast\}$ given by
	\[
		\pi_{F \setminus \{\ast\}} \otimes \id \: : \: X_J \otimes X_\ast \longrightarrow X_{F \setminus \{\ast\}} \otimes X_\ast
	\]
	for $\ast \in F$, and just
	\[
		\pi_F \otimes \discard{X_\ast} : X_J \otimes X_\ast \longrightarrow X_F 
	\]
	in case $\ast \not \in F$. It is straightforward to check that these morphisms indeed make $X_J \otimes X_\ast$ carry the universal property of the infinite tensor product $X_{J \sqcup \{\ast\}}$ if and only if the functor $- \otimes X_\ast$ preserves the defining cofiltered limit of the infinite tensor product $X_J$. It is also easy to see that this new cofiltered limit is again preserved by every functor $- \otimes Y$, based on the assumption that the defining cofiltered limit of the original infinite tensor product is preserved by $- \otimes (X_\ast \otimes Y)$ as well.

	In summary, the preservation condition in this definition amounts to the requirement that the canonical comparison morphism~\eqref{comparison} must be an isomorphism, as one would expect intuitively from a good notion of infinite tensor product. 
\end{remark}

\begin{example}
	\label{infprods_cring}
	The dual definition of infinite tensor products of algebraic structures as \emph{filtered colimits} of finite tensor products is well-known in the literature, e.g.~in the case of C*-algebras~\cite[p.~315]{blackadar}, although we do not know whether the preservation condition has been made explicit before. For example in the case where our category is $\CRing_+\op$, we recover the usual folklore definition of an infinite tensor product of rings $\bigotimes_{i \in J} R_i$ in terms of formal sums of elementary tensors, where an elementary tensor is a family of elements $(r_i)_{i \in J}$ such that all but finitely many are equal to the respective unit. The necessary preservation condition is easily seen to hold. Note that we do not yet consider the multiplication on $\bigotimes_{i \in J} R_i$, since \Cref{semicartesian_infproduct} is not yet concerned with the comonoid structures on the objects. We will get to this in the next section.

	Intuitively, our definition matches up with the known algebraic ones under the categorical duality of algebra and geometry, where our \Cref{semicartesian_infproduct} is on the geometrical side of the duality.
\end{example}

\begin{remark}
	In those semicartesian monoidal categories $\cC$ that are of interest to us, morphisms $I \to X$ play the role of \emph{probability measures} on $X$. Thus applying the defining universal property of an infinite tensor product with respect to maps out of $I$ implements the \emph{Kolmogorov extension theorem} in $\cC$: probability measures on an infinite product $X_J$ are in bijection with consistent families of probability measures on the finite products $X_F$ for $F \subseteq J$. The spirit of \Cref{semicartesian_infproduct} is thus such that it turns the Kolmogorov extension theorem into a definition.
\end{remark}

@@

	\label{infprods_stoch}
	We now consider infinite tensor products in $\Stoch$, where we would like infinite tensor products to be given by the corresponding infinite products of measurable spaces in the usual sense. Since the Kolmogorov extension theorem does not hold for general (even merely countable) products of measurable spaces~\cite{AJ}, this is not the case without further additional assumptions on the measurable spaces involved. Thus we do not know whether $\Stoch$ has infinite tensor products, although we suspect that it does not; but even if it does, they are in general not the ones that one will care about for the purposes of probability theory.

	However, the situation improves for countable products in $\BorelStoch$, for which the Kolmogorov extension theorem indeed holds~\cite[Theorem~14.35]{klenke}. In other words, if $\left( (X_j, \Sigma_j) \right)_{j \in \bN}$ is a sequence of standard Borel spaces, then the cartesian product $X_\bN = \prod_{j \in \bN} X_j$, carrying the product $\sigma$-algebra $\Sigma_\bN$, satisfies the universal property of an infinite tensor product with respect to maps out of $I$. We now prove that this implies the universal property in general.

	Thus suppose that
	\[
		\Big( g_F : (A,\Sigma_A) \longrightarrow (X_F, \Sigma_F) \Big)_{F \subseteq J \text{ finite}}
	\]
	is a family of Markov kernels satisfying the compatibility condition $\pi_{F',F} \circ g_{F'} = g_F$ for all finite $F \subseteq F' \subseteq J$. Then for every $a \in A$, the probability measures $g_F(-|a) : \Sigma_F \to [0,1]$ are a compatible family to which the Kolmogorov extension theorem in the form of~\cite[Theorem~14.35]{klenke} applies, and we obtain a unique probability measure $g_J(-|a) : \Sigma_J \to [0,1]$ which has the original $g_F(-|a)$ as its finite marginals. It remains to be shown that for every $S \in \Sigma_J$, the map $a \mapsto g_J(S|a)$ is measurable. Since limits of pointwise convergent sequences of measurable real-valued functions are again measurable, the set of $S$ for which this measurability holds is closed under countable disjoint union, and it is clearly closed under complements. We therefore have a $\lambda$-system. Since the map is measurable by assumption whenever $S$ is a measurable cylinder set, and the cylinder sets form a $\pi$-system, the $\pi$-$\lambda$-theorem implies that $a \mapsto g_J(S|a)$ is measurable for all $S$ in the $\sigma$-algebra generated by the measurable cylinder sets, which is exactly the product $\sigma$-algebra $\Sigma_J$.

@@

\begin{definition}
    \label{defn_kolmogorov_ext}
    Let $\cC$ be a Markov category and $(X_i)_{i \in J}$ a family of objects. We say that an infinite tensor product $X_J = \bigotimes_{i \in J} X_i$ is a \emph{Kolmogorov product} if the finite marginalization morphisms $\pi_F : X_J \to X_F$ are deterministic.
\end{definition}

In $\BorelStoch$, the infinite tensor products constructed in \Cref{infprods_stoch} are Kolmogorov products. In $\CRing_+\op$, the infinite tensor products from \Cref{infprods_cring} are Kolmogorov products as well, provided that one equips $\bigotimes_i R_i$ with the tensor product ring structure, since this one is the only one which makes the canonical inclusions $R_j \to \bigotimes_i R_i$ into ring homomorphisms. These examples illustrate the following important point. In some Markov categories, such as $\CRing_+\op$, not all isomorphisms are deterministic, and there may even be isomorphic objects which have no deterministic isomorphism~\cite[Remark~10.9]{markov_cats}. In these categories, whether a given infinite tensor product is a Kolmogorov product or not generally depends on the specific choice of that object. In particular, there may be infinite tensor products which are not Kolmogorov products. For example in $\CRing_+\op$, we may take the tensor product of abelian groups $\bigotimes_i R_i$ and equip it with a different multiplication than the tensor product of rings one (as long as it has the same unit).
